Lattice Boltzmann simulations of the dynamics of thin liquid films \\ \large{Gitter Boltzmann Simulationen der Dynamik von d\"unnen Fl\"ussigkeitsfilmen}

### 2024-08-01 chapters/Theory.tex
Updates from Overleaf

For the sake of clarity however here it is assumed that both $\mu$ and $\lambda$ are constants.
Inclusion of pressure to $\hat{\sigma}$ yields the total stress tensor $\tilde{\sigma}$. 
Due to symmetries of the problem there are various constraints on this tensor.
A fluid element can be rotated similar to a solid body, but it can also be stretched. 
In contrast to a solid body a fluid element is allowed to deform.
Working out these constraints we find that the pressure has to be a diagonal term and thus
\begin{equation}\label{eq:total_stress}
    \tilde{\sigma}_{\alpha\beta} = - p \delta_{\alpha\beta} + \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}}.
\end{equation}
Now to get from the Euler equation (Eq.~(\ref{eq:navier_stokes_2})) to the Navier-Stoke we add the viscous stress tensor and rearrange terms,
\begin{equation}\label{eq:navier_stokes_3}
    \partial_t(\rho\vec{u}) + \vec{\nabla}\cdot(\rho\vec{u}\vec{u}) = - \vec{\nabla} p + \vec{\nabla}\left[\mu(\vec{\nabla}\vec{u} + \vec{\nabla}\vec{u}^T) + \left(\eta-\frac{2\mu}{3}\right)\vec{\nabla}\cdot\vec{u}\right] + \vec{F},
\end{equation}
where we have used the differential form of Eq.~(\ref{eq:total_stress}) and $\lambda = \eta - 2/3\mu$.

@@

Scaling analysis and renormalization are key concepts in modern physics~\cite{yakhotRenormalizationGroupAnalysis1986, IntroductionQuantumField2018}.
One of the key concepts of this idea are symmetries, and therefore transformations under which a given state does not change.
Fluid dynamics to some extent admits a conformal symmetry.
The governing equations describe the flow through a nanotube as well as the circular currents in oceans~\cite{secchiMassiveRadiusdependentFlow2016, marcheDerivationNewTwodimensional2007}.
Some features however such as the drag on an airfoil do not simply scale with the size of the system.
Therefore while a miniaturised model may create enough lift force, simply scaling up the size is not sufficient to ensure enough lift.  
In the following it will be helpful to introduce dimensionless numbers~\cite{ruzickaDimensionlessNumbers2008,neversFluidMechanicsChemical2019}.
These numbers have a long history in fluid dynamics.
The Reynolds number for example is a measure if a flow is turbulent or not, it depends only on proportions and not on an exact experiment.
With the knowledge of these number comparing experiments is straightforward, even if the experiments differ in many aspects.
The two dimensionless numbers of interest for the shallow water theory are the Reynolds number and the Froude number
\begin{equation}\label{eq:Re_and_Fr}
    Re = \frac{\rho L v_0}{\mu},\quad Fr = \frac{v_0}{\sqrt{g L}},
\end{equation}
where $L$ is a characteristic length scale of the problem, e.g., the mean water depth, $v_0$ and $g$ are a characteristic velocity and the gravitational acceleration ($g = 9.81 m/s^2$), respectively.
A characteristic velocity can be for example the inflow velocity from a boundary, or the speed of a certain wave type such as a tsunami.
Both numbers measure the magnitude of certain effects, in the case of the Reynolds number the inertial forces are compared against viscosity.
The Froude number is a measure to compare kinetic ($v_0$) with potential energy ($g L$).

In the next step both $Fr$ and $Re$ are used to make Eq.~(\ref{eq:navier_stokes_fin}) dimensionless.
For convenience the vector equation is split into two respective equations with gravity being the only force present.
Writing the full equation with all components yields
\begin{align}\label{eq:2d_RNS}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_z u_x &= -\partial_x p + \frac{1}{Re}\Delta u_x ,\\
    \partial_t u_z + u_z\partial_x u_x + u_z\partial_z u_z &= -\partial_z p - \frac{1}{Fr^2} + \frac{1}{Re}\Delta u_z ,
\end{align}
with the two dimensional Laplacian $\Delta = (\partial_x^2 + \partial_z^2)$.
Using long wave scaling, Eq.~(\ref{eq:long_wave_scaling}) and the rescaled length and time variables (e.g., $\tilde{x}$),
\begin{align}\label{eq:2d_RNS_eps}
    \varepsilon(\partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x \partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z \partial_{\tilde{z}} \tilde{u}_x ) &= -\varepsilon\partial_{\tilde{x}} \tilde{p} + \frac{1}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_x + \partial_{\tilde{z}}^2 \tilde{u}_x ) ,\\
    \varepsilon^2(\partial_{\tilde{t}} \tilde{u}_z + \tilde{u}_z\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_z) &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_z + \partial_{\tilde{z}}^2 \tilde{u}_z ) .
\end{align}
We now assume that terms with higher orders in $\varepsilon$ are negligible and therefore $O(\varepsilon^2) \rightarrow 0$, collecting only terms with up to $O(\varepsilon)$
\begin{align}\label{eq:2d_RNS_linear}
    \partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_x &= -\partial_{\tilde{x}} \tilde{p} + \frac{\partial_{\tilde{z}}^2\tilde{u}_x}{\varepsilon Re}, \\
    0 &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re} .
\end{align}
Interestingly in the first of the two above equations the term associated to the horizontal viscosity is $O(\varepsilon^2)$ and can be dropped.
On the other hand with the assumption of the long wave scaling the second of the above equations yields the hydrostatic pressure, taking only $O(1)$ terms
\begin{equation}\label{eq:hydro_static}
    \partial_{\tilde{z}}\tilde{p} = -\frac{1}{Fr^2},
\end{equation}
which can be readily integrated, given $Fr$ is a constant, using the upper bound at the liquid air interface $\eta(x)$ 
\begin{equation}\label{eq:hydro_static_int}
    \tilde{p} = \frac{1}{Fr^2}(\tilde{\eta} - \tilde{z}).   
\end{equation}
This implies that $\partial_{\tilde{x}}p$ does not depend on $\tilde{z}$ and therefore states that the pressure gradient is conserved along z.

Summarizing the impact of the approximation, on the continuity equation Eq.~(\ref{eq:rescaled_cont}) and Eqs.~(\ref{eq:2d_RNS_linear}-\ref{eq:hydro_static}) and dropping the tilde we get 
\begin{align}
    \partial_x u_x + \partial_z u_z &= 0, \label{eq:cont_sw_euler}\\
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x &= -\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}, \label{eq:mom_sw_euler}\\
    \partial_z p &= -\frac{1}{Fr^2} \left(+\frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re}\right), \label{eq:hydrostat_sw_euler}
\end{align}
with the addition of boundary conditions at the liquid air interface ($z = \eta$)
\begin{align}
    \partial_t \eta + u_x\partial_x\eta - u_z &= 0, \label{eq:sw_boundaries_top_1}\\
    p &= 0, \label{eq:sw_boundaries_top_2}\\
    \partial_z u_x &= 0, \label{eq:sw_boundaries_top_3}
\end{align}
as well as the no slip condition at the liquid solid interface ($z = b$)
\begin{align}
    u_x = 0, \label{eq:no-slip_sw1}\\
    u_z = 0. \label{eq:no-slip_sw2}
\end{align}
One interesting limit is the case when $\varepsilon Re \rightarrow \infty$.
In this limit both the continuity equation as well as the boundary conditions are untouched.
The only part that is modified is the momentum equation Eq.~(\ref{eq:mom_sw_euler}),
\begin{equation}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x = -\partial_x p,
\end{equation}
which describes the hydrostatic Euler equations~\cite{brenierGeneralizedSolutionsHydrostatic2008}.

@@

Performing the horizontal integration to the depth averaged momentum equation, see Eq.~(\ref{eq:mom_sw_euler}) we have
\begin{align}\label{eq:mom_shallow_water_int}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z. 
\end{align}
The two terms on the left hand side can be integrated straight away. 
We know that the pressure gradient $\partial_x p$ is conserved along the vertical dimension. 
The other term ($\propto\partial^2 u$) has a similar shape as the viscous term in Eq.~(\ref{eq:navier_stokes_4}). 
Performing the integral therefore yields
\begin{equation}\label{eq:sw_mom_left}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p + \frac{\partial_z u_x|_{z = \eta} - \partial_z u_x|_{z=b}}{\varepsilon Re},
with the boundary condition that fluid does not accelerate into the air phase, see Eq.~(\ref{eq:sw_boundaries_top_3}), the left hand side reduces to
\begin{equation}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p - \frac{\partial_z u_x|_{z=b}}{\varepsilon Re}.
The right hand side of Eq.~(\ref{eq:mom_shallow_water_int}) reads
\begin{align}\label{eq:mom_shallow_rhs}
    \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x\right)\diff z \nonumber\\
    &+ \int_b^{\eta}\left(u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z ,
\end{align}
where the integral is split into two separate integrals for convenience.
The time dependent part can be easily integrated using Eq.(~\ref{eq:int_leibniz})
\begin{equation}\label{eq:first_term_rhs_mom_sw}
    \int_b^{\eta}\left(\partial_t u_x\right)\diff z = \partial_t \int_b^{\eta} u_x \diff z - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b} = \partial_t(hU) - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b}. 
The other integral is somewhat more tricky.
First, the $u_z\partial_z u_x$ term is integrated by parts
\begin{equation}
    \int_b^{\eta}u_z(\partial_{z} u_x)\diff z = [u_x u_z]_b^{\eta} - \int_b^{\eta} (\partial_z u_z) u_x \diff z.
We use this intermediate result, Eq.~(\ref{eq:cont_sw_euler}) and the integration rule of Leibnitz to perform the integral of the second line of Eq.(\ref{eq:mom_shallow_rhs}) 
\begin{align}
    \int_b^{\eta}(u_x\partial_x u_x &+ u_z\partial_{z} u_x )\diff z = 2\int_b^{\eta} u_x\partial_x u_x \diff z + [u_x u_z]_b^{\eta},\nonumber \\
     &= \partial_x\left(\int_b^{\eta}u_x^2 \diff z\right) + \partial_t\eta u_x|_{z=\eta} + u_x(u_x\partial_x b - u_z)|_{z=b},
\end{align}
where the same line of arguments as in Eq.~(\ref{eq:vert_int_cont_3}) has been used.
Applying the boundary condition to this equation and using the definition for height and discharge we have
\begin{equation}\label{eq:sw_mom_final}
    \partial_t(hU) + \partial_x(\beta hU^2) = -h\partial_x p - \tau_b,
where two new variables appear.
The first one is the Boussinesq coefficient $\beta$ which is defined by the equation
\begin{equation}\label{eq:boussinesq}
    \int_b^{\eta} u_x^2 \diff z = \beta h U^2,
and the second one is the friction or the shear stress at the bottom of the fluid column
\begin{equation}\label{eq:sw_bot_friction}
    \tau_b = \frac{\partial_z u_x|_{z=b}}{\varepsilon Re}. 
Both $\beta$ as well as $\tau_b$ do depend on $u$. 
Therefore to know their exact contribution one needs to know about the flow.
Without the concrete knowledge of $u$ there is however one observation concerning Eq.~(\ref{eq:boussinesq})
\begin{equation}\label{eq:beta_lims}
    \beta = 1 + \frac{1}{h}\int_b^{\eta}\left(1 - \frac{u}{U}\right)^2 \diff z,
and therefore $\beta \geq 1$.
To understand the implications of these terms we discuss an instructive example.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In second publication we show how this translates into a numerical solver.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{graphics/contact_line.pdf}
    \caption{A liquid droplet placed on a smooth substrate.
    The three phase contact line which is the interface between the vapor, the solid and liquid phase is displayed as a thick black line.}
    \label{fig:contact_line_drop}
\end{figure}
Similar to the shallow water system, the thin film equation only considers the dynamics of a thin liquid \textit{layer}. 
In principle this layer can be unsupported, it can as well be placed on top of a yet undefined surface or substrate.
Within this thesis we consider exclusively supported thin films on substrates.
The defining equation that describes the dynamics of a thin liquid film can be written as~\cite{thielePatternedDepositionMoving2014, oronLongscaleEvolutionThin1997, crasterDynamicsStabilityThin2009, bonnWettingSpreading2009} 
\begin{equation}\label{eq:simple_thin_film}
    \partial_t h = \nabla\cdot\left(M(h)\nabla p\right).
\end{equation}
On the left hand side we have the change in thickness ($\partial_t h$), which can only be achieved by applying a pressure gradient ($\nabla p$).
In Eq.~(\ref{eq:simple_thin_film}) two unknowns are introduced.
The first one is the mobility $M(h)$ which determines the exact velocity boundary condition for $h=0$.
Using the no-slip boundary condition~\cite{beckerThinfilmEquationRecent2005}, see Eq.~(\ref{eq:no-slip_sw1}), we have 
\begin{equation}\label{eq:no-slip-mobility}
    M(h) = \frac{h^3}{3\mu},
\end{equation}
with $\mu$ being the (dynamic) viscosity. 
Therefore the mobility is inversely proportional to the viscosity.
Of course the velocity boundary condition can be relaxed which is often desired e.g., for contact line movement.

@@

Before we start the derivation we rescale length and velocity components to have dimensionless quantities.
We use a similar scaling as in Eq.~(\ref{eq:thickness_length}) and get 
\begin{align}\label{eq:thin_film_scaling}
    \tilde{x} &= \frac{x}{L},\quad \tilde{y} = \frac{y}{L},\quad \tilde{z} = \frac{z}{H},\nonumber \\
    \tilde{u}_x &= \frac{u_x}{U},\quad \tilde{u}_y = \frac{u_y}{U},\quad \tilde{u}_z = \frac{u_z}{V},\nonumber \\
    \tilde{t} &= \frac{L}{U} t,\quad \tilde{p} = \frac{H^2 p}{\mu U L}, 
\end{align}
where $U$ and $V$ are characteristic velocities and $L$ and $H$ are characteristic length scales in horizontal and vertical dimension.
The scaling factor in front of the pressure can be calculated assuming stationary unidirectional flow with $\mathbf{u} = (u(z), 0, 0)$.
It is this scaling factor ($\mu U L /H^2$) that requires that inertial effects are small and $Re<1$.
In many applications the small Reynolds number limit is a good approximation, however for e.g., spin coating inertia is not negligible.
To account for inertial effects the scaling, Eq.~(\ref{eq:thin_film_scaling}), must be modified. 
For example the pressure scaling would change to $p = \rho U^2\tilde{p}$~\cite{crasterDynamicsStabilityThin2009}.

>>>>

For the sake of clarity however here it is assumed that both $\mu$ and $\lambda$ are constants.
Inclusion of pressure to $\hat{\sigma}$ yields the total stress tensor $\tilde{\sigma}$. 
Due to symmetries of the problem there are various constraints on this tensor.
A fluid element can be rotated similar to a solid body, but it can also be stretched. 
In contrast to a rigid body, an idealized model of a solid body, a fluid element is allowed to deform.
Working out these constraints we find that the pressure has to be a diagonal term and thus
\begin{equation}\label{eq:total_stress}
    \tilde{\sigma}_{\alpha\beta} = - p \delta_{\alpha\beta} + \mu\left(\frac{\partial u_{\alpha}}{\partial x_{\beta}} + \frac{\partial u_{\beta}}{\partial x_{\alpha}}\right) + \lambda\delta_{\alpha\beta}\frac{\partial u_{\gamma}}{\partial x_{\gamma}}.
\end{equation}
Now to get from the Euler equation (Eq.~(\ref{eq:navier_stokes_2})) to the Navier-Stoke we add the viscous stress tensor and rearrange terms,
\begin{equation}\label{eq:navier_stokes_3}
    \partial_t(\rho\vec{u}) + \vec{\nabla}\cdot(\rho\vec{u}\vec{u}) = - \vec{\nabla} p + \vec{\nabla}\left[\mu(\vec{\nabla}\vec{u} + \vec{\nabla}\vec{u}^T) + \left(\eta-\frac{2\mu}{3}\right)\vec{\nabla}\cdot\vec{u}\right] + \vec{F},
\end{equation}
where we have used the differential form of Eq.~(\ref{eq:total_stress}) and $\lambda = \eta - 2/3\mu$.

@@

Scaling analysis and renormalization are key concepts in modern physics~\cite{yakhotRenormalizationGroupAnalysis1986, IntroductionQuantumField2018}.
One of the key concepts of this idea are symmetries, and therefore transformations under which a given state does not change.
Fluid dynamics to some extent admits a conformal symmetry.
The governing equations describe the flow through a nanotube as well as the circular currents in oceans~\cite{secchiMassiveRadiusdependentFlow2016, marcheDerivationNewTwodimensional2007}.
Some features however such as the drag on an airfoil do not simply scale with the size of the system.
Therefore while a miniaturised model may create enough lift force, simply scaling up the size is not sufficient to ensure enough lift.  
In the following it will be helpful to introduce dimensionless numbers~\cite{ruzickaDimensionlessNumbers2008,neversFluidMechanicsChemical2019}.
These numbers have a long history in fluid dynamics.
%The Reynolds number for example is a measure if a flow is turbulent or not, it depends only on proportions and not on an exact experiment.
The Reynolds number ($Re$) is such a dimensionless number and is calculated as the ratio of characteristic inertial and viscous forces.
It is often used to distinguish between laminar and turbulent flows, however there is not a singular value that sets a sharp boundary between the two regimes as it is very specific to the geometry of the problem.
% More often than not it depends on the specific experiment.
That said, with the knowledge of these number comparing experiments is straightforward, even if the experiments differ in many aspects.
The two dimensionless numbers of interest for the shallow water theory are the Reynolds number and the Froude number
\begin{equation}\label{eq:Re_and_Fr}
    Re = \frac{\rho L v_0}{\mu},\quad Fr = \frac{v_0}{\sqrt{g L}},
\end{equation}
where $L$ is a characteristic length scale of the problem, e.g., the mean water depth, $v_0$ and $g$ are a characteristic velocity and the gravitational acceleration ($g = 9.81 m/s^2$), respectively.
A characteristic velocity can be for example the inflow velocity from a boundary, or the speed of a certain wave type such as a tsunami.
Both numbers measure the magnitude of certain effects.
The Froude number is a measure to compare kinetic ($v_0$) with potential energy ($g L$).

In the next step both $Fr$ and $Re$ are used to make Eq.~(\ref{eq:navier_stokes_fin}) dimensionless.
For convenience the vector equation is split into two respective equations where the gravity contribution can be understood as the arbitrary $\vec{F}$ term on the right.
Writing the full equation with all components yields
\begin{align}\label{eq:2d_RNS}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_z u_x &= -\partial_x p + \frac{1}{Re}\Delta u_x ,\\
    \partial_t u_z + u_z\partial_x u_x + u_z\partial_z u_z &= -\partial_z p - \frac{1}{Fr^2} + \frac{1}{Re}\Delta u_z ,
\end{align}
with the two dimensional Laplacian $\Delta = (\partial_x^2 + \partial_z^2)$.
Using long wave scaling, Eq.~(\ref{eq:long_wave_scaling}) and the rescaled length and time variables (e.g., $\tilde{x}$),
\begin{align}\label{eq:2d_RNS_eps}
    \varepsilon(\partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x \partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z \partial_{\tilde{z}} \tilde{u}_x ) &= -\varepsilon\partial_{\tilde{x}} \tilde{p} + \frac{1}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_x + \partial_{\tilde{z}}^2 \tilde{u}_x ) ,\\
    \varepsilon^2(\partial_{\tilde{t}} \tilde{u}_z + \tilde{u}_z\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_z) &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon}{Re}(\varepsilon^2 \partial_{\tilde{x}}^2 \tilde{u}_z + \partial_{\tilde{z}}^2 \tilde{u}_z ) .
\end{align}
We now assume that terms with higher orders in $\varepsilon$ are negligible and therefore $O(\varepsilon^2) \rightarrow 0$, collecting only terms with up to $O(\varepsilon)$
\begin{align}\label{eq:2d_RNS_linear}
    \partial_{\tilde{t}} \tilde{u}_x + \tilde{u}_x\partial_{\tilde{x}} \tilde{u}_x + \tilde{u}_z\partial_{\tilde{z}} \tilde{u}_x &= -\partial_{\tilde{x}} \tilde{p} + \frac{\partial_{\tilde{z}}^2\tilde{u}_x}{\varepsilon Re}, \\
    0 &= -\partial_{\tilde{z}}\tilde{p} - \frac{1}{Fr^2} + \frac{\varepsilon\partial_{\tilde{z}}^2 \tilde{u}_z}{Re} .
\end{align}
Interestingly in the first of the two above equations the term associated to the horizontal viscosity is $O(\varepsilon^2)$ and can be dropped.
On the other hand with the assumption of the long wave scaling the second of the above equations yields the hydrostatic pressure, taking only $O(1)$ terms
\begin{equation}\label{eq:hydro_static}
    \partial_{\tilde{z}}\tilde{p} = -\frac{1}{Fr^2},
\end{equation}
which can be readily integrated, given $Fr$ is a constant, using the upper bound at the liquid air interface $\eta(x)$ 
\begin{equation}\label{eq:hydro_static_int}
    \tilde{p} = \frac{1}{Fr^2}(\tilde{\eta} - \tilde{z}).   
\end{equation}
This implies that $\partial_{\tilde{x}}p$ does not depend on $\tilde{z}$ and therefore states that the pressure gradient is conserved along z.

Summarizing the impact of the approximation, on the continuity equation Eq.~(\ref{eq:rescaled_cont}) and Eqs.~(\ref{eq:2d_RNS_linear}-\ref{eq:hydro_static}) and dropping the tilde we get 
\begin{align}
    \partial_x u_x + \partial_z u_z &= 0, \label{eq:cont_sw_euler}\\
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x &= -\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}, \label{eq:mom_sw_euler}\\
    \partial_z p &= -\frac{1}{Fr^2} \left(+\frac{\varepsilon\partial_{\tilde{z}}^2 u_z}{Re}\right), \label{eq:hydrostat_sw_euler}
\end{align}
with the addition of boundary conditions at the liquid air interface ($z = \eta$)
\begin{align}
    \partial_t \eta + u_x\partial_x\eta - u_z &= 0, \label{eq:sw_boundaries_top_1}\\
    p &= 0, \label{eq:sw_boundaries_top_2}\\
    \partial_z u_x &= 0, \label{eq:sw_boundaries_top_3}
\end{align}
as well as the no slip condition at the liquid solid interface ($z = b$)
\begin{align}
    u_x = 0, \label{eq:no-slip_sw1}\\
    u_z = 0. \label{eq:no-slip_sw2}
\end{align}
One interesting limit is the case when $\varepsilon Re \rightarrow \infty$.
In this limit both the continuity equation as well as the boundary conditions are untouched.
The only part that is modified is the momentum equation Eq.~(\ref{eq:mom_sw_euler}),
\begin{equation}
    \partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x = -\partial_x p,
\end{equation}
which describes the hydrostatic Euler equations~\cite{brenierGeneralizedSolutionsHydrostatic2008}.

@@

Performing the horizontal integration to the depth averaged momentum equation, see Eq.~(\ref{eq:mom_sw_euler}) we have
\begin{align}\label{eq:mom_shallow_water_int}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z. 
\end{align}
The two terms on the left hand side can be integrated straight away. 
We know that the pressure gradient $\partial_x p$ is conserved along the vertical dimension. 
The other term ($\propto\partial^2 u$) has a similar shape as the viscous term in Eq.~(\ref{eq:navier_stokes_4}). 
Performing the integral therefore yields
\begin{equation}\label{eq:sw_mom_left}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p + \frac{\partial_z u_x|_{z = \eta} - \partial_z u_x|_{z=b}}{\varepsilon Re},
\end{equation}
with the boundary condition that fluid does not accelerate into the air phase, the left hand side reduces to
\begin{equation}
    \int_b^{\eta}\left(-\partial_x p + \frac{\partial_z^2 u_x}{\varepsilon Re}\right)\diff z = -h\partial_x p - \frac{\partial_z u_x|_{z=b}}{\varepsilon Re}.
\end{equation}
The right hand side of Eq.~(\ref{eq:mom_shallow_water_int}) reads
\begin{align}\label{eq:mom_shallow_rhs}
    \int_b^{\eta}\left(\partial_t u_x + u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z &= \int_b^{\eta}\left(\partial_t u_x\right)\diff z \nonumber\\
    &+ \int_b^{\eta}\left(u_x\partial_x u_x + u_z\partial_{z} u_x\right)\diff z ,
\end{align}
where the integral is split into two separate integrals for convenience.
The time dependent part can be easily integrated using Eq.(~\ref{eq:int_leibniz})
\begin{equation}\label{eq:first_term_rhs_mom_sw}
    \int_b^{\eta}\left(\partial_t u_x\right)\diff z = \partial_t \int_b^{\eta} u_x \diff z - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b} = \partial_t(hU) - \partial_t\eta u_x|_{z=\eta} - \partial_t b u_x|_{z=b}. 
\end{equation}
The other integral is somewhat more tricky.
First, the $u_z\partial_z u_x$ term is integrated by parts
\begin{equation}
    \int_b^{\eta}u_z(\partial_{z} u_x)\diff z = [u_x u_z]_b^{\eta} - \int_b^{\eta} (\partial_z u_z) u_x \diff z.
\end{equation}
We use this intermediate result, Eq.~(\ref{eq:cont_sw_euler}) and the integration rule of Leibnitz to perform the integral of the second line of Eq.(\ref{eq:mom_shallow_rhs}) 
\begin{align}
    \int_b^{\eta}(u_x\partial_x u_x &+ u_z\partial_{z} u_x )\diff z = 2\int_b^{\eta} u_x\partial_x u_x \diff z + [u_x u_z]_b^{\eta},\nonumber \\
     &= \partial_x\left(\int_b^{\eta}u_x^2 \diff z\right) + \partial_t\eta u_x|_{z=\eta} + u_x(u_x\partial_x b - u_z)|_{z=b},
\end{align}
where the same line of arguments as in Eq.~(\ref{eq:vert_int_cont_3}) has been used.
Applying the boundary condition to this equation and using the definition for height and discharge we have
\begin{equation}\label{eq:sw_mom_final}
    \partial_t(hU) + \partial_x(\beta hU^2) = -h\partial_x p - \tau_b,
\end{equation}
where two new variables appear.
The first one is the Boussinesq coefficient $\beta$ which is defined by the equation
\begin{equation}\label{eq:boussinesq}
    \int_b^{\eta} u_x^2 \diff z = \beta h U^2,
\end{equation}
and the second one is the friction or the shear stress at the bottom of the fluid column
\begin{equation}\label{eq:sw_bot_friction}
    \tau_b = \frac{\partial_z u_x|_{z=b}}{\varepsilon Re}. 
\end{equation}
Both $\beta$ as well as $\tau_b$ do depend on $u$. 
Therefore to know their exact contribution one needs to know about the flow.
Without the concrete knowledge of $u$ there is however one observation concerning Eq.~(\ref{eq:boussinesq})
\begin{equation}\label{eq:beta_lims}
    \beta = 1 + \frac{1}{h}\int_b^{\eta}\left(1 - \frac{u}{U}\right)^2 \diff z,
\end{equation}
and therefore $\beta \geq 1$.
To understand the implications of these terms we discuss an instructive example.

@@

To close this section we quickly collect the equations that form the shallow water system, see Eqs.~(\ref{eq:shallow_water_cont_true}, \ref{eq:sw_mom_final}).
The generalization of these equations towards a second horizontal dimension is straightforward and can be written as~\cite{salmonLatticeBoltzmannMethod1999, dellarNonhydrodynamicModesPriori2002, thommesLatticeBoltzmannMethods2007}
\begin{align}    
        \partial_t h &+ \nabla \cdot (h\mathbf{u}) = 0, \label{eq:shallow_water_01}\\
        \partial_t (h\mathbf{u}) + \nabla&\cdot (p \mathbb{1} + h\mathbf{u}\mathbf{u} - S) = 0, \label{eq:shallow_water_02} 
\end{align}
where $\mathbf{u} = (u_x, u_y)$ and the only unknown $\mathcal{S}$ is a yet undefined source term. 
The system Eqs.~(\ref{eq:shallow_water_01}-\ref{eq:shallow_water_02}) concludes the shallow water model. 
In the second publication we show how this translates into a numerical solver.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.95\textwidth]{graphics/contact_line.pdf}
    \caption{A liquid droplet placed on a smooth substrate.
    The three phase contact line which is the interface between the vapor, the solid and liquid phase is displayed as a thick black line.}
    \label{fig:contact_line_drop}
\end{figure}
Similar to the shallow water system, the thin film equation only considers the dynamics of a thin liquid \textit{layer}. 
In principle this layer can be unsupported, it can as well be placed on top of a yet undefined surface or substrate.
Within this thesis we consider exclusively supported thin films on substrates.
The defining equation that describes the dynamics of a thin liquid film can be written as~\cite{thielePatternedDepositionMoving2014, oronLongscaleEvolutionThin1997, crasterDynamicsStabilityThin2009, bonnWettingSpreading2009} 
\begin{equation}\label{eq:simple_thin_film}
    \partial_t h = \nabla\cdot\left[M(h)\nabla p\right].
\end{equation}
On the left hand side we have the change in thickness ($\partial_t h$), which can only be achieved by applying a pressure gradient ($\nabla p$).
In Eq.~(\ref{eq:simple_thin_film}) two unknowns are introduced.
The first one is the mobility $M(h)$ which determines the exact velocity boundary condition for $h=0$.
Using the no-slip boundary condition~\cite{beckerThinfilmEquationRecent2005}, see Eq.~(\ref{eq:no-slip_sw1}), we have 
\begin{equation}\label{eq:no-slip-mobility}
    M(h) = \frac{h^3}{3\mu},
\end{equation}
with $\mu$ being the (dynamic) viscosity. 
Therefore the mobility is inversely proportional to the viscosity.
Of course the velocity boundary condition can be relaxed which is often desired e.g., for contact line movement.

@@

Before we start the derivation we rescale length and velocity components to have dimensionless quantities.
We use a similar scaling as in Eq.~(\ref{eq:thickness_length}) and get 
\begin{align}\label{eq:thin_film_scaling}
    \tilde{x} &= \frac{x}{L},\quad \tilde{y} = \frac{y}{L},\quad \tilde{z} = \frac{z}{H},\nonumber \\
    \tilde{u}_x &= \frac{u_x}{U},\quad \tilde{u}_y = \frac{u_y}{U},\quad \tilde{u}_z = \frac{u_z}{V},\nonumber \\
    \tilde{t} &= \frac{U}{L} t,\quad \tilde{p} = \frac{H^2 p}{\mu U L}, 
\end{align}
where $U$ and $V$ are characteristic velocities and $L$ and $H$ are characteristic length scales in horizontal and vertical dimension.
The scaling factor in front of the pressure can be calculated assuming stationary unidirectional flow with $\mathbf{u} = (u(z), 0, 0)$.
It is this scaling factor ($\mu U L /H^2$) that requires that inertial effects are small and $Re<1$.
In many applications the small Reynolds number limit is a good approximation, however for e.g., spin coating inertia is not negligible.
To account for inertial effects the scaling, Eq.~(\ref{eq:thin_film_scaling}), must be modified. 
For example the pressure scaling would change to $p = \rho U^2\tilde{p}$~\cite{crasterDynamicsStabilityThin2009}.
